The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2012-12-13 script/en/chapters/cryptomethods.tex
CT script: Minor changes

% HACK to fix warning "destination with the same identifier .. has already been used, ...":
\makeatletter \renewcommand{\thepage}{~\csname @arabic\endcsname \c@page~} \makeatother
\hypertarget{Kapitel_1}{}
\chapter{Security Definitions and Encryption Procedures}
\label{Label_Kapitel_1}
(Bernhard Esslinger, Joerg-Cornelius Schneider, May 1999; Updates Dec. 2001, Feb. 2003,
June 2005, July 2007, Jan. 2010, Nov. 2012)

@@

% --------------------------------------------------------------------------
\section[Asymmetric encryption]
{Asymmetric encryption\footnotemark}
  \footnotetext{%
    With CrypTool 1\index{CrypTool 1} you can execute RSA encryption
    and decryption (using the menu path {\bf Crypt \textbackslash{} Asymmetric}).
    In both cases you must select a RSA key pair. Only in the case of decryption
    the secret RSA key is necessary -- so here you are asked to enter the PIN.\\
    With CrypTool 2\index{CrypTool 2} you can also perform asymmetic methods
    (using the templates under {\bf Cryptography \textbackslash{} Modern}).\\
    JCrypTool\index{JCrypTool} offers asymmetic methods like RSA both within
    the {\bf Visuals} menu of the Default Perspective as well as within the
    Functional Perspective.
  }
\nopagebreak
In the case of {\em asymmetric} encryption \index{Encryption!asymmetric} each
subscriber has a personal pair of keys consisting of a {\em secret}
\index{Key!secret} key and a {\em public} key\index{Key!public}. The public
key, as its name implies, is made public, e.g. in a key directory on the
Internet.\par \vskip + 3pt

@@

Such a procedure can be demonstrated using a series of thief-proof letter boxes.
If I have composed a message, I then look for the letter box of the recipient
and post the letter through it. After that, I can no longer read or change the
message myself, because only the legitimate recipient has the key for the
letter box.\par \vskip + 3pt

The advantage of asymmetric procedures is the easy \index{Key management} key management. Let's look again at a network with $n$
subscribers. In order to ensure that each subscriber can establish
an encrypted connection to each other subscriber, each subscriber
must possess a pair of keys. We therefore need $2n$ keys or $n$
pairs of keys. Furthermore, no secure channel is needed before
messages are transmitted, because all the information required in
order to communicate confidentially can be sent openly. In
this case, you simply have to pay attention to the accuracy
(integrity and authenticity) \index{Authenticity} of the public
key. Disadvantage: Pure asymmetric procedures take a lot longer to
perform than symmetric ones.\par \vskip + 3pt

@@

% --------------------------------------------------------------------------
% \newpage
\section[Hybrid procedures]{Hybrid procedures\footnotemark}
\footnotetext{%
   Within CT1\index{CrypTool 1} you can find this technique using the menu
   path {\bf Crypt \textbackslash{} Hybrid}: 
   There you can follow the single steps and its dependencies with concrete
   numbers. The variant with RSA as the asymmetric algorithm is graphically
   visualized; the variant with ECC uses the standard dialogs. In both cases
   AES is used as the symmetric algorithm.\index{AES}\\
   JCrypTool\index{JCrypTool} offers hybrid methods like ECIES within
   the Functional Perspective under {\bf Algorithms \textbackslash{} Hybrid Ciphers}.
}
\label{CM_Hybrid-procedures}
\index{Hybrid procedure}

@@

In this case the bulk data is encrypted using symmetric procedures: The key
used for this is a secret session key\index{Session key} generated by the
sender randomly\footnote{%
   An important part of cryptographically secure techniques is to generate 
   random numbers. Within CT1\index{CrypTool 1} you can check out
   different random number generators using the menu path
   {\bf Indiv. Procedures \textbackslash{} Generate Random Numbers}. 
   Using the menu path {\bf Analysis \textbackslash{} Analyze Randomness}
   you can apply different test methods for random data to binary documents. \\
   CT1 concentrates on cryptographically strong {\bf pseudo} random number
   generators. Only the integrated Secude generator involves a "pure" random source.\\
   JCrypTool\index{JCrypTool} offers {\bf pseudo} random number generators both
   within the Default Perspective in the menu {\bf Algorithms \textbackslash{}
   Random Number Generator} as well as within the Functional Perspective.
}\index{Random}
that is only used for this message.

@@

Compared to public-key ciphers based on mathematics like RSA, the structure of AES,
and most other modern symmetric ciphers, is very complex and cannot be explained
as easily as RSA.\index{AES}

So some simplified variants of modern symmetric ciphers like DES, IDEA or AES
were developed for educational purposes in order to allow beginners to perform
encryption and decryption by hand and gain a better understanding of how the
algorithms work in detail.
These simplified variants also help to understand and apply the according
cryptanalysis methods.

One of these variants is e.g. S-AES (Simplified-AES) by Prof. Ed Schaefer
and his students \cite{cm:Musa-etal2003}%
\footnote{
    See the two articles with the same title
    ``Devising a Better Way to Teach and Learn the Advanced Encryption Standard''
    at the university's news at\\
    - \url{http://math.scu.edu/~eschaefe/getfile.pdf} \\
    - \url{http://www.scu.edu/cas/research/cryptography.cfm}
}.
Another one is Mini-AES \cite{cm:Phan2002}
(see chapter~\ref{CM_Sage_Mini-AES} ``\nameref{CM_Sage_Mini-AES}'')%
\index{DES}\index{DES!SDES}\index{IDEA}:
\begin{itemize}

@@

@@

\bibitem[Haan2008]{cm:Haan2008} \index{Haan 2008} 
       Kristian Laurent Haan, \\
       {\em Advanced Encryption Standard (AES)},\\
       2008, (German article)\\
       \url{http://www.codeplanet.eu/tutorials/cpp/51-advanced-encryption-standard.html}

\bibitem[Hoffman2006]{cm:Hoffman2006} \index{Hoffman 2006} 
       Nick Hoffman, \\
       {\em A SIMPLIFIED IDEA ALGORITHM},\\
       2006 \\
       \url{http://www.nku.edu/~christensen/simplified%20IDEA%20algorithm.pdf}

@@

\bibitem[Schmeh2003]{cm:Schmeh2003}  \index{Schmeh 2003}
       Klaus Schmeh, \\
       {\em Cryptography and Public Key Infrastructures on the Internet},\\ 
       John Wiley \& Sons Ltd., Chichester 2003. \\
       An up-to-date, easy to read book, which also considers practical
       problems such as standardization or real existing software.
       In German the 4th edition was published in 2009.

@@

% --------------------------------------------------------------------------
\newpage
\section*{Web links}
\addcontentsline{toc}{section}{Web links}

@@

  \hypertarget{CM_HT_Weblink_Rijndael-Cryptosystem}{}
  \item AES or Rijndael cryptosystem \\
        \url{http://www.cryptosystem.net/aes} \\
	\url{http://www.minrank.org/aes/}

  \item AES discussion groups at NIST \\
	\url{http://aes.nist.gov/aes}

  \item distributed.net: ``RC5-64 has been solved'' \\
        \url{http://distributed.net/pressroom/news-20020926.html}

  \item RSA: ``The RSA Secret Key Challenge'' \\
      \url{http://www.rsasecurity.com/rsalabs/challenges/secretkey/index.html}

  \item RSA: ``DES Challenge'' \\
        \url{http://www.rsasecurity.com/rsalabs/challenges/des3/index.html}

@@

>>>>

% HACK to fix warning "destination with the same identifier .. has already been used, ...":
\makeatletter \renewcommand{\thepage}{~\csname @arabic\endcsname \c@page~} \makeatother
\hypertarget{Kapitel_1}{}
\chapter{Security Definitions and Encryption Procedures}
\label{Label_Kapitel_1}
(Bernhard Esslinger, Joerg-Cornelius Schneider, May 1999; Updates Dec. 2001, Feb. 2003,
June 2005, July 2007, Jan. 2010, Dec. 2012)

@@

% --------------------------------------------------------------------------
\section[Asymmetric encryption]
{Asymmetric encryption\footnotemark}
  \footnotetext{%
    With CT1\index{CrypTool 1} you can execute RSA encryption and decryption
    (using the menu path {\bf Crypt \textbackslash{} Asymmetric}).
    In both cases you must select a RSA key pair. Only in the case of decryption
    the secret RSA key is necessary -- so here you are asked to enter the PIN.\\
    With CT2\index{CrypTool 2} you can also perform asymmetic methods
    (using the templates under {\bf Cryptography \textbackslash{} Modern}).\\
    JCT\index{JCrypTool} offers asymmetic methods like RSA both within
    the {\bf Visuals} menu of the Default Perspective as well as within the
    Algorithm Perspective.
  }
\nopagebreak
In the case of {\em asymmetric} encryption \index{Encryption!asymmetric} each
subscriber has a personal pair of keys consisting of a {\em secret}
\index{Key!secret} key and a {\em public} key\index{Key!public}. The public
key, as its name implies, is made public, e.g. in a key directory on the
Internet (this kind of ``bill-board'' is also called just directory or
public key ring).\par \vskip + 3pt

@@

Such a procedure can be demonstrated using a series of thief-proof letter boxes.
If I have composed a message, I then look for the letter box\index{Letter box}
of the recipient and post the letter through it. After that, I can no longer
read or change the message myself, because only the legitimate recipient has
the key for the letter box.\par \vskip + 3pt

The advantage of asymmetric procedures is the easier \index{Key management}
key management. Let's look again at a network with $n$
subscribers. In order to ensure that each subscriber can establish
an encrypted connection to each other subscriber, each subscriber
must possess a pair of keys. We therefore need $2n$ keys or $n$
pairs of keys. Furthermore, no secure channel is needed before
messages are transmitted, because all the information required in
order to communicate confidentially can be sent openly. In
this case, you simply%
\footnote{%
That this is also not trivial is explained e.g. in xxxx % [[[KES-Article-SharedPrimes__und__anderes-Chapter-hier]]]
. Besides the requirements
for the key generation it has to be considered hat nowadays also (public-key)
infrastructures are targets of attacks.
}%
have to pay attention to the accuracy
(integrity and authenticity) \index{Authenticity} of the public
key. Disadvantage: Pure asymmetric procedures take a lot longer to
perform than symmetric ones.\par \vskip + 3pt

@@

% --------------------------------------------------------------------------
% \newpage
\section[Hybrid procedures]{Hybrid procedures\footnotemark}
\footnotetext{%
   Within CT1\index{CrypTool 1} you can find this technique using the menu
   path {\bf Crypt \textbackslash{} Hybrid}: 
   There you can follow the single steps and its dependencies with concrete
   numbers. The variant with RSA as the asymmetric algorithm is graphically
   visualized; the variant with ECC uses the standard dialogs. In both cases
   AES is used as the symmetric algorithm.\index{AES}\\
   JCT\index{JCrypTool} offers hybrid methods like ECIES within
   the Algorithm  Perspective under {\bf Algorithms \textbackslash{} Hybrid Ciphers}.
}
\label{CM_Hybrid-procedures}
\index{Hybrid procedure}

@@

In this case the bulk data is encrypted using symmetric procedures: The key
used for this is a secret session key\index{Session key} generated by the
sender randomly\footnote{%
   An important part of cryptographically secure techniques is to generate 
   random numbers. Within CT1\index{CrypTool 1} you can check out
   different random number generators using the menu path
   {\bf Indiv. Procedures \textbackslash{} Generate Random Numbers}. 
   Using the menu path {\bf Analysis \textbackslash{} Analyze Randomness}
   you can apply different test methods for random data to binary documents. \\
   CT1 concentrates on cryptographically strong {\bf pseudo} random number
   generators. ``Real'' random sources are only included via calling the
   integrated Secude generator.\\
   JCT\index{JCrypTool} offers {\bf pseudo} random number generators both
   within the Default Perspective in the menu {\bf Algorithms \textbackslash{}
   Random Number Generator} as well as within the Algorithm Perspective.
}\index{Random}
that is only used for this message.

@@

Compared to public-key ciphers based on mathematics like RSA, the structure of AES\index{AES} and most other modern symmetric ciphers (like DES, IDEA or Present), is very complex and cannot be explained as easily as RSA\index{RSA}.

So simplified variants of modern symmetric ciphers were developed for
educational purposes in order to allow beginners to perform
encryption and decryption by hand and gain a better understanding of how the
algorithms work in detail.
These simplified variants also help to understand and apply the according
cryptanalysis methods.

One of these variants is e.g. S-AES (Simplified-AES) by Prof. Ed Schaefer
and his students \cite{cm:Musa-etal2003}.%
\footnote{
    See the two articles with the same title
    ``Devising a Better Way to Teach and Learn the Advanced Encryption Standard''
    at the university's news at\\
    - \url{http://math.scu.edu/~eschaefe/getfile.pdf} \\
    - \url{http://www.scu.edu/cas/research/cryptography.cfm}
}
Another one is Mini-AES \cite{cm:Phan2002}
(see chapter~\ref{CM_Sage_Mini-AES} ``\nameref{CM_Sage_Mini-AES}'')%
\index{DES}\index{DES!SDES}\index{IDEA}:
\begin{itemize}

@@

% ---------------------------------------------------------------------------
\subsection{Further symmetric crypto algorithms in Sage}
\label{CM_Sage_SymCryptoAlg}

The Sage Reference v5.4.1%
\footnote{
See \url{http://www.sagemath.org/doc/reference/sage/crypto/}
}
lists the following further ciphers:\\ 
- Linear feedback shift register (LFSR),\\
- ShrinkingGeneratorCryptosystem,\\
- Blum-Blum-Shub (BBS) pseudorandom bit generator,\\
- Lattice related functions.

@@

\bibitem[Haan2008]{cm:Haan2008} \index{Haan 2008} 
       Kristian Laurent Haan, \\
       {\em Advanced Encryption Standard (AES)},
       2008, (German article)\\
       \url{http://www.codeplanet.eu/tutorials/cpp/51-advanced-encryption-standard.html}

\bibitem[Hoffman2006]{cm:Hoffman2006} \index{Hoffman 2006} 
       Nick Hoffman, \\
       {\em A SIMPLIFIED IDEA ALGORITHM},
       2006 \\
       \url{http://www.nku.edu/~christensen/simplified%20IDEA%20algorithm.pdf}

@@

\bibitem[Schmeh2003]{cm:Schmeh2003}  \index{Schmeh 2003}
       Klaus Schmeh, \\
       {\em Cryptography and Public Key Infrastructures on the Internet},\\ 
       John Wiley \& Sons Ltd., Chichester 2003. \\
       An easy to read book, which also considers practical
       problems such as standardization or real existing software.
       In German, the 5th edition was published in 2012.

@@

% --------------------------------------------------------------------------
\newpage
% \section*{Web links}\addcontentsline{toc}{section}{Web links}
\chapter*{Web Links}\addcontentsline{toc}{section}{Web Links}

@@

  \hypertarget{CM_HT_Weblink_Rijndael-Cryptosystem}{}
  \item AES or Rijndael cryptosystem \\
        \url{http://www.cryptosystem.net/aes}

  \item AES discussion groups at NIST \\
	\url{http://csrc.nist.gov/archive/aes/}

  \item distributed.net: ``RC5-64 has been solved'' \\
        \url{http://www.distributed.net/Pressroom_press-rc5-64}

  \item RSA: ``The RSA Secret Key Challenge'' \\
        \url{http://www.rsa.com/rsalabs/node.asp?id=2100}

  \item RSA: ``DES Challenge'' \\
        \url{http://www.rsa.com/rsalabs/node.asp?id=2108}

@@

All links have been confirmed at December 12, 2012.
